\documentclass[mast]{lucky}



\title{Perspectives}
\author{Dennis Chen}
\date{CQV}

\begin{document}

\maketitle

\input{CQV-Perspectives}

\pagebreak

\section{Problems}
\minpt{40}

\psetquote{I won't ever allow you to do anything to abandon these possibilities and choose death!}{Fullmetal Alchemist: Brotherhood}

     \begin{prob}[AMC 8 2011/6]{2}
In a town of $351$ adults, every adult owns a car, motorcycle, or both. If $331$ adults own cars and $45$ adults own motorcycles, how many of the car owners do not own a motorcycle?
\end{prob}

     \begin{prob}[]{2}
How many integers from $1$ to $100$ (inclusive) are multiples of $2$ or $3?$
\end{prob}

     \begin{prob}[AMC 10A 2018/11]{3}
When 7 fair standard 6-sided dice are thrown, the probability that the sum of the numbers on the top faces is 10 can be written as\[\frac{n}{6^7},\]where $n$ is a positive integer. What is $n$?
\end{prob}

     \begin{prob}[]{2}
There are $3$ distinct six-sided dice, one red, white, and blue. How many ways can the sum of the $15$ faces showing on the three die equal $56,$ if each die orientation is only considered unique if the sum of its faces that are showing are unique?
\end{prob}
    
     \begin{prob}[AMC 10B 2017/13]{2}
There are $20$ students participating in an after-school program offering classes in yoga, bridge, and painting. Each student must take at least one of these three classes, but may take two or all three. There are $10$ students taking yoga, $13$ taking bridge, and $9$ taking painting. There are $9$ students taking at least two classes. How many students are taking all three classes?
\end{prob}
    
     \begin{req}[AMC 10B 2020/23]{3}
Square $ABCD$ in the coordinate plane has vertices at the points $A(1,1), B(-1,1), C(-1,-1),$ and $D(1,-1).$ Consider the following four transformations:
    
\begin{itemize}
    \Item $L,$ a rotation of $90^{\circ}$ counterclockwise around the origin;

    \Item $R,$ a rotation of $90^{\circ}$ clockwise around the origin;
    
    \Item $H,$ a reflection across the $x$-axis; and

    \Item $V,$ a reflection across the $y$-axis.
\end{itemize}

Each of these transformations maps the squares onto itself, but the positions of the labeled vertices will change. For example, applying $R$ and then $V$ would send the vertex $A$ at $(1,1)$ to $(-1,-1)$ and would send the vertex $B$ at $(-1,1)$ to itself. How many sequences of $20$ transformations chosen from $\{L, R, H, V\}$ will send all of the labeled vertices back to their original positions? (For example, $R, R, V, H$ is one sequence of $4$ transformations that will send the vertices back to their original positions.
\end{req}

\begin{req}[AIME I 2020/7]{3}
A club consisting of $11$ men and $12$ women needs to choose a committee from among its members so that the number of women on the committee is one more than the number of men on the committee. The committee could have as few as $1$ member or as many as $23$ members. Let $N$ be the number of such committees that can be formed. Find the sum of the prime numbers that divide $N.$
\end{req}
    
\begin{prob}[]{3}
We have $7$ balls each of different colors (red, orange, yellow, green, blue, indigo, violet) and $3$ boxes each of different shapes (tetrahedron, cube, dodecahedron). How many ways are there to place these $7$ balls into the $3$ boxes such that each box contains at least $1$ ball?
\end{prob}
    
\begin{prob}[AIME II 2009/6]{3}
Let $m$ be the number of five-element subsets that can be chosen from the set of the first $14$ natural numbers so that at least two of the five numbers are consecutive. Find the remainder when $m$ is divided by $1000$. 
\end{prob}  
    
\begin{prob}[AIME II 2002/9]{4}
Let $\mathcal{S}$ be the set $\lbrace1,2,3,\ldots,10\rbrace.$ Let $n$ be the number of sets of two non-empty disjoint subsets of $\mathcal{S}$. (Disjoint sets are defined as sets that have no common elements.) Find the remainder obtained when $n$ is divided by $1000$.
\end{prob}
    
     \begin{prob}[Mildorf AIME]{4}
Let $N$ denote the number of $7$ digit positive integers have the property that their digits are in increasing order. Determine the remainder obtained when $N$ is divided by $1000$. (Repeated digits are allowed.)
\end{prob}

     \begin{prob}[AIME I 2020/9]{4}
Let $S$ be the set of positive integer divisors of $20^9.$ Three numbers are chosen independently and at random with replacement from the set $S$ and labeled $a_1,a_2,$ and $a_3$ in the order they are chosen. The probability that both $a_1$ divides $a_2$ and $a_2$ divides $a_3$ is $\tfrac{m}{n},$ where $m$ and $n$ are relatively prime positive integers. Find $m.$
\end{prob}
    
     \begin{prob}[AIME II 2013/9]{6}
A $7\times 1$ board is completely covered by $m\times 1$ tiles without overlap; each tile may cover any number of consecutive squares, and each tile lies completely on the board. Each tile is either red, blue, or green. Let $N$ be the number of tilings of the $7\times 1$ board in which all three colors are used at least once. For example, a $1\times 1$ red tile followed by a $2\times 1$ green tile, a $1\times 1$ green tile, a $2\times 1$ blue tile, and a $1\times 1$ green tile is a valid tiling. Note that if the $2\times 1$ blue tile is replaced by two $1\times 1$ blue tiles, this results in a different tiling. Find the remainder when $N$ is divided by $1000$.
\end{prob}
    
     \begin{req}[AIME I 2015/12]{6}
Consider all 1000-element subsets of the set $\{1, 2, 3,\ldots,2015\}.$ From each such subset choose the least element. The arithmetic mean of all of these least elements is $\frac{p}{q}$, where $p$ and $q$ are relatively prime positive integers. Find $p + q$.
\end{req}
    
     \begin{prob}[AIME 1986/13]{9}
In a sequence of coin tosses, one can keep a record of instances in which a tail is immediately followed by a head, a head is immediately followed by a head, and etc. We denote these by TH, HH, and etc. For example, in the sequence TTTHHTHTTTHHTTH of 15 coin tosses we observe that there are two HH, three HT, four TH, and five TT subsequences. How many different sequences of 15 coin tosses will contain exactly two HH, three HT, four TH, and five TT subsequences?
\end{prob}
    
     \begin{prob}[CMIMC 2018]{13}
Compute the number of rearrangements $a_1,a_2,\ldots,a_{2018}$ of the sequence $1,2,\ldots,2018$ such that $a_k>k$ for exactly one value of $k.$
\end{prob}

\end{document}