\documentclass{article}

\usepackage[mast]{lucky}

\title{Perspectives}
\author{Dennis Chen}
\date{CQV}

\begin{document}

\maketitle

\input{CQV-Perspectives}

\pagebreak

\section{Problems}

\minpt{42}

\psetquote{Human nature is a sullen, obstinate, unreasonable brute; but it always has its own way with all of us.}{A Fool's Errand}

     \begin{prob}[]{2}
Katie is teaching a class of thirty pupils, of which fourteen are girls. She knows that there are twenty-two pupils who are right-handed. What is the minimum amount of girls that are right-handed?
\end{prob}
    
     \begin{prob}[]{2}
In a higher secondary examination $80\%$ of the examinees have passed in English and $85\%$ in Mathematics, while $75\%$ passed in both English and Mathematics. If $45$ candidates failed in both the subjects, find the total number of candidates.
\end{prob}

     \begin{prob}[]{2}
We have $10$ students taking Spanish, $12$ students taking French, and $16$ people taking German. We also have $4$ people taking Spanish and French, $5$ people taking French and German, and $7$ people taking German and Spanish. Then we have $3$ very brave souls who are taking, Spanish, French, and German! How many students are taking at least one language?
\end{prob}
    
     \begin{prob}[AIME I 2002/1]{2}
Many states use a sequence of three letters followed by a sequence of three digits as their standard license-plate pattern. Given that each three-letter three-digit arrangement is equally likely, the probability that such a license plate will contain at least one palindrome (a three-letter arrangement or a three-digit arrangement that reads the same left-to-right as it does right-to-left) is $\dfrac{m}{n}$, where $m$ and $n$ are relatively prime positive integers. Find $m+n.$
\end{prob}

\begin{prob}[MATHCOUNTS State Sprint 2020/27]{2}
How many of the first $2019$ positive integers have no odd single-digit prime factors?
\end{prob}
    
     \begin{prob}[AMC 10A 2020/23]{3}
Let $T$ be the triangle in the coordinate plane with vertices $(0,0), (4,0),$ and $(0,3).$ Consider the following five isometries (rigid transformations) of the plane: rotations of $90^{\circ}, 180^{\circ},$ and $270^{\circ}$ counterclockwise around the origin, reflection across the $x$-axis, and reflection across the $y$-axis. How many of the $125$ sequences of three of these transformations (not necessarily distinct) will return $T$ to its original position? (For example, a $180^{\circ}$ rotation, followed by a reflection across the $x$-axis, followed by a reflection across the $y$-axis will return $T$ to its original position, but a $90^{\circ}$ rotation, followed by a reflection across the $x$-axis, followed by another reflection across the $x$-axis will not return $T$ to its original position.)
\end{prob}
    
    \begin{prob}[Scrabbler AMC 10]{3}
Robert writes all positive divisors of the number $216$ on separate slips of paper, then places the slips into a hat. He randomly selects three slips from the hat, with replacement. What is the probability that the product of the numbers on the three slips Robert selects is a divisor of $216$?
\end{prob}
    
    \begin{prob}[AIME 1993/8]{4}
Let $S$ be a set with six elements. In how many different ways can one select two not necessarily distinct subsets of $S$ so that the union of the two subsets is $S$? The order of selection does not matter; for example, the pair of subsets $\{a, c\}$, $\{b, c, d, e, f\}$ represents the same selection as the pair $\{b, c, d, e, f\}$, $\{a, c\}$.
\end{prob}
    
    \begin{req}[AIME 1999/13]{4}
Forty teams play a tournament in which every team plays every other team exactly once. No ties occur, and each team has a $50 \%$ chance of winning any game it plays. The probability that no two teams win the same number of games is $\frac mn,$ where $m_{}$ and $n_{}$ are relatively prime positive integers. Find $\log_2 n.$
\end{req}
    
    \begin{prob}[AIME II 2017/9]{4}
A special deck of cards contains $49$ cards, each labeled with a number from $1$ to $7$ and colored with one of seven colors. Each number-color combination appears on exactly one card. Sharon will select a set of eight cards from the deck at random. Given that she gets at least one card of each color and at least one card with each number, the probability that Sharon can discard one of her cards and still have at least one card of each color and at least one card with each number is $\frac{p}{q}$, where $p$ and $q$ are relatively prime positive integers. Find $p+q$.
\end{prob}
    
    \begin{prob}[AIME I 2017/7]{6}
For nonnegative integers $a$ and $b$ with $a + b \leq 6$, let $T(a, b) = \binom{6}{a} \binom{6}{b} \binom{6}{a + b}$. Let $S$ denote the sum of all $T(a, b)$, where $a$ and $b$ are nonnegative integers with $a + b \leq 6$. Find the remainder when $S$ is divided by $1000$.
\end{prob}
    
    \begin{req}[AIME 1992/12]{6}
n a game of Chomp, two players alternately take bites from a 5-by-7 grid of unit squares. To take a bite, a player chooses one of the remaining squares, then removes ("eats") all squares in the quadrant defined by the left edge (extended upward) and the lower edge (extended rightward) of the chosen square. For example, the bite determined by the shaded square in the diagram would remove the shaded square and the four squares marked by $\times.$ (The squares with two or more dotted edges have been removed form the original board in previous moves.)

\begin{center}
\begin{asy}
defaultpen(linewidth(0.7));
fill((2,2)--(2,3)--(3,3)--(3,2)--cycle, mediumgray);
int[] array={5, 5, 5, 4, 2, 2, 2, 0};
pair[] ex = {(2,3), (2,4), (3,2), (3,3)};
draw((3,5)--(7,5)^^(4,4)--(7,4)^^(4,3)--(7,3), linetype("3 3"));
draw((4,4)--(4,5)^^(5,2)--(5,5)^^(6,2)--(6,5)^^(7,2)--(7,5), linetype("3 3"));
int i, j;
for(i=0; i<7; i=i+1) {
for(j=0; j<array[i]; j=j+1) {
draw((i,j+1)--(i,j)--(i+1,j));
}
draw((i,array[i])--(i+1,array[i]));
if(array[i]>array[i+1]) {
draw((i+1,array[i])--(i+1,array[i+1]));
}}
for(i=0; i<4; i=i+1) {
draw(ex[i]--(ex[i].x+1, ex[i].y+1), linewidth(1.2));
draw((ex[i].x+1, ex[i].y)--(ex[i].x, ex[i].y+1), linewidth(1.2));
}
\end{asy}
\end{center}

The object of the game is to make one's opponent take the last bite. The diagram shows one of the many subsets of the set of 35 unit squares that can occur during the game of Chomp. How many different subsets are there in all? Include the full board and empty board in your count. 
\end{req}

\vspace{0.2cm}

\begin{prob}[IMO 1981/2]{6}
Let $1\leq r\leq n.$ We consider all $r$-element subsets of $\{1,\ldots,n\}.$ Each of them has a minimum. Prove that the average of these minima is $\frac{n+1}{r+1}.$
\end{prob}

\begin{prob}[CIME II 2021/6]{6}
A rat on the coordinate plane starts from the point $(0,0).$ Every second, if the rat is currently on the point $(x,y),$ it moves to one of the points $(x+1,y)$, $(x,y+1)$, $(x+1,y-1)$, $(x-1,y+1).$

\noindent The rat may not visit the same point more than once, and every point $(x,y)$ that rat visits staisfies $0\leq x, y\leq 4.$ Let $N$ be the number of ways for the rat to move from $(0,0)$ to $(4,4).$ Find the number of positive divisors of $N.$
\end{prob}

\begin{prob}[USAMO 1990/4]{9}
Find, with proof, the number of positive integers whose base-$n$ representation consists of distinct digits with the property that, except for the leftmost digit, every digit differs by $\pm 1$ from some digit further to the left. (Your answer should be an explicit function of $n$ in simplest form.)
\end{prob}

\begin{req}[AMC 12A 2021/15]{9}
A choir direction must select a group of singers from among his $6$ tenors and $8$ basses. The only requirements are that the difference between the number of tenors and basses must be a multiple of $4$, and the group must have at least one singer. Let $N$ be the number of groups that can be selected. What is the remainder when $N$ is divided by $100$?
\end{req}

    
    \begin{prob}[Scrabbler AMC 12]{13}
A positive integer is monotone if its digits, when read left-to-right, are either in strictly increasing or strictly decreasing order. For example, 7, 540, and 24578 are monotone numbers, while 0 and 9986 are not. How many monotone numbers leave a remainder of 1 when divided by 3?
\end{prob}
    
    \begin{prob}[HMMT Guts 2020/15]{13}
You have six blocks in a row, labeled 1 through 6. Call two blocks $x \le y$
    connected when, for all $x \le z \le y$, block $z$ has not been removed. While there is still at least one block remaining, you choose a remaining block uniformly at random and remove it. The cost of this operation is the the number of blocks that are connected to the block being removed, including itself. Compute the expected total cost of removing all the blocks.
\end{prob}
\end{document}


