\documentclass[mast]{lucky}



\title{Radical Axes}
\author{Dennis Chen}
\date{GRU}

\begin{document}

\maketitle

\input{GRU-Radical}

\pagebreak
\section{Problems}
\minpt{TBD}

\begin{prob}[]{1}
Consider a point $P$ with power $36$ respective to a circle with center $O.$ If $PO=10,$ find the radius of the circle.
\end{prob}

\begin{prob}[]{2}
Consider a coordinate plane with two circles tangent to the $x$ axis at $X,Y,$ respectively. If the circles intersect at $P,Q,$ and $XY=8,$ is it possible for $P$ to lie on $y=3$ and $Q$ to lie on $y=12?$
\end{prob}

\begin{prob}[APMO 2020/1]{3}
Let $\Gamma$ be the circumcircle of $\triangle ABC$. Let $D$ be a point on the side $BC$. The tangent to $\Gamma$ at $A$ intersects the parallel line to $BA$ through $D$ at point $E$. The segment $CE$ intersects $\Gamma$ again at $F$. Suppose $B$, $D$, $F$, $E$ are concyclic. Prove that $AC$, $BF$, $DE$ are concurrent.

\end{prob}

\begin{prob}[EGMO 2019/4]{3}
Let $ABC$ be a triangle with incentre $I$. The circle through $B$ tangent to $AI$ at $I$ meets side $AB$ again at $P$. The circle through $C$ tangent to $AI$ at $I$ meets side $AC$ again at $Q$. Prove that $PQ$ is tangent to the incircle of $ABC.$
\end{prob}

\begin{prob}[IMO 2004/1]{4}
Let $ABC$ be an acute-angled triangle with $AB\neq AC$. The circle with diameter $BC$ intersects the sides $AB$ and $AC$ at $M$ and $N$ respectively. Denote by $O$ the midpoint of the side $BC$. The bisectors of the angles $\angle BAC$ and $\angle MON$ intersect at $R$. Prove that the circumcircles of the triangles $BMR$ and $CNR$ have a common point lying on the side $BC$.
\end{prob}

\begin{prob}{4}
Let $P_1P_2P_3\dots P_{2n}$ be a regular $2n$-gon, and let $X$ be a point. Prove \[(P_1P_{n+1}X), (P_2P_{n+2}X), (P_3P_{n+3}X),\dots (P_nP_{2n}X)\] are concurrent.
\end{prob}

\begin{prob}[ISL 2017/G1]{6}
Let $ABCDE$ be a convex pentagon such that $AB=BC=CD$, $\angle{EAB}=\angle{BCD}$, and $\angle{EDC}=\angle{CBA}$. Prove that the perpendicular line from $E$ to $BC$ and the line segments $AC$ and $BD$ are concurrent.
\end{prob}

\begin{prob}[Swiss TST 2018/3]{6}
Let $ABC$ be a triangle with $\angle  ABC \ne \angle  BCA$. The inscribed circle $k$ of the triangle $ABC$ is tangent to the sides $BC, CA,AB$ at points $D, E , F$ respectively. The segment $AD$ intersects $k$ again at $P$. Let $Q$ be the point of intersection of $EF$ with the line perpendicular to $AD$ passing through $P$. Let $X,Y$ be the points of intersection of $AQ$ with $DE,DF$ respectively. Prove that $A$ is the midpoint of the segment $XY$.
\end{prob}

\begin{prob}[ISL 2017 G5]{9}
Let $ABCC_1B_1A_1$ be a convex hexagon such that $AB=BC$, and suppose that the line segments $AA_1, BB_1$, and $CC_1$ have the same perpendicular bisector. Let the diagonals $AC_1$ and $A_1C$ meet at $D$, and denote by $\omega$ the circle $ABC$. Let $\omega$ intersect the circle $A_1BC_1$ again at $E \neq B$. Prove that the lines $BB_1$ and $DE$ intersect on $\omega$.
\end{prob}

\end{document}